---
editor: 
  markdown: 
    wrap: 72
---
# Getting to know your Data with R

### Installing required packages

```{r}
library(pacman)
```

### Loading data

```{r}
p_load(dlookr,
       DMwR2, # Data Mining with R functions
       GGally, # Pair-wise plots using ggplot2
       Hmisc, # Data analysis 
       palmerpenguins, # Alternative to the Iris dataset
       tidyverse) # Data wrangling, manipulation, visualization
```

```{r}
algae |> glimpse()
```

## Central Tendency

#### Mean
```{r}
algae$a1 |>  mean()
```
#### Median
```{r}
algae$a1 |> median()
```
#### Mode
```{r}
Mode <- function(x, na.rm=FALSE){
  if(na.rm) x<-x[!is.na(x)]
  ux <- unique (x)
  return (ux[which.max(tabulate(match(x, ux)))])
}

algae$a2 |> Mode()
```

## DMwR centralValue() function

#### Numerical variable
```{r}
algae$a1 |> centralValue()
```

#### Nominal variable
```{r}
algae$speed |> centralValue()
```


## Statistics of spread (variation)

#### Variance
```{r}
algae$a1 |> var()
```
#### Standard Deviation
```{r}
algae$a1 |> sd()
```
#### Range
```{r}
algae$a1 |> range()
```
#### Minimum Value
```{r}
algae$a1 |> min()
```
#### Maximum Value
```{r}
algae$a1 |> max()
```
#### Interquartile Range
```{r}
algae$a1 |> IQR()
```
#### Quantile
```{r}
algae$a1 |> quantile()
```
#### Specifying specific quantile
```{r}
algae$a1 |> quantile(probs = c(0.2, 0.8))
```
## Missing values

```{r}
library(purrr)
```

### Compute the total number of NA values in the dataset

```{r}
nas <- algae %>% purrr::map_dbl(~sum(is.na(.))) %>% sum()
cat("The dataset contains", nas, "NA values. \n")
```
### Compute the number of incomplete rows in the dataset

```{r}
incomplete_rows <- algae %>%
  summarise_all(~!complete.cases(.)) %>%
  nrow() 
cat("The dataset contains", incomplete_rows, "(out of", nrow(algae),") incomplete rows. \n")
```

## Summary of the dataset

#### base r's summary()
```{r}
algae |> summary()
```

#### Hmisc's describe()
```{r}
data("penguins")
penguins |>
  Hmisc::describe()
```

### dlookr's describe()
```{r}
penguins |>
  dlookr::describe()
```

## Summaries on a subset of data

```{r}
algae |>
  summarise(avgNO3 = mean(NO3, na.rm=TRUE),
            medA1 = median(a1))
algae |>
  select(mxPH:Cl) |>
  summarise_all(list(mean, median), na.rm = TRUE)
algae |>
  select(a1:a7) |>
  summarise_all(funs(var))
algae |>
  select(a1:a7) |>
  summarise_all(c("min", "max"))
algae |>
  group_by(season, size) |>
  summarise(nObs = n(), mA7 = median(a7))
penguins |> 
  group_by(species) |>
  summarise(var = var(bill_length_mm, na.rm = TRUE))
```

## Aggregating data

```{r}
penguins |>
  group_by(species) |>
  reframe(var = quantile(bill_length_mm, na.rm = TRUE))
penguins |>
  group_by(species) |>
  dlookr::describe(bill_length_mm)
```

